---
title: "Better Management of Bicycle Fleet - BikeShare"
output: 
  powerpoint_presentation:
      reference_doc: Bikes_Template.pptx
df_print: paged
always_allow_html: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
rm(list=ls(all=TRUE))

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(tidymodels)) install.packages("tidymodels", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repunios = "http://cran.us.r-project.org")
if(!require(patchwork)) install.packages("patchwork", repos = "http://cran.us.r-project.org")
if(!require(gridtext)) install.packages("gridtext", repos = "http://cran.us.r-project.org")
if(!require(grid)) install.packages("grid", repos = "http://cran.us.r-project.org")
if(!require(magick)) install.packages("magick", repos = "http://cran.us.r-project.org")
if(!require(gtools)) install.packages("gtools", repos = "http://cran.us.r-project.org")
if(!require(ggmap)) install.packages("gtools", repos = "http://cran.us.r-project.org")
if(!require(mltools)) install.packages("mltools", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("knitr", repos = "http://cran.us.r-project.org")
if(!require(officer)) install.packages("knitr", repos = "http://cran.us.r-project.org")


library(magick)
library(tidyverse)
library(tidymodels)
library(ggthemes)
library(patchwork)
library(grid)
library(gridtext)
library(rvest)
library(RColorBrewer)
library(lubridate)
library(flextable)
library(officer)

seasons <- tibble(season=c("DJF","MAM","JJA","SON"),
                  season_name=c("winter","spring","summer","autumn"))
    
```


```{r load_data, echo=FALSE, include=FALSE}
if(file.exists("trip_data.RDS")){
  trip_data <- readRDS("trip_data.RDS")
  
}else{
  library(aws.s3) 
  #filesat https://s3.amazonaws.com/capitalbikeshare-data/index.html"
  bucket_id <- 'capitalbikeshare-data'
  bucket <- get_bucket(bucket_id,max = Inf)
  
  keys <- c()
  for(i in seq(1:length(bucket))){
    keys[i] <- bucket[[i]]$Key
  }
  #ignore 2020 
  keys <- c(keys[grepl('2018',keys)],keys[grepl('2019',keys)],keys[grepl('2017',keys)])
  
  save_object(keys[1], file = keys[1], bucket = bucket_id)
  unzip(keys[1],exdir = "./zip")
  file.remove(keys[1])
  files <- list.files("./zip")
  files <- files[!grepl('MACOSX',files)]
  #first file
  
  trip_data <- read_csv(paste("./zip/",files[1],sep=""))
  
  if(length(files)>1){
    for(j in 2:length(files)){
      trip_data_j <- read_csv(paste("./zip/",files[j],sep=""))
      trip_data <- rbind(trip_data,trip_data_j)
      rm(trip_data_j)
    }
    file.remove(paste("./zip/",files,sep=""))
  }
  
  
  ###other files
  
  if(length(keys)>1){
    for(i in 1:length(keys)){
      
      save_object(keys[i], file = keys[i], bucket = bucket_id)
      unzip(keys[i],exdir = "./zip")
      file.remove(keys[i])
      files <- list.files("./zip") 
      files <- files[!grepl('MACOSX',files)]
      
      for(j in 1:length(files)){
        trip_data_j <- read_csv(paste("./zip/",files[j],sep=""))
        trip_data <- rbind(trip_data,trip_data_j)
        rm(trip_data_j)
      }
      file.remove(paste("./zip/",files,sep=""))
      
    }
  }
  
  saveRDS(trip_data,"trip_data.RDS")
}


if(file.exists("weather_data.RDS")){
 weather_data <- readRDS("weather_data.RDS")
}else{
  files <- list.files("./weather/")
  weather_data <- read.csv(paste("./weather/",files[1],sep=""))

  if(length(files)>1){
    for(j in 2:length(files)){
      weather_data_j <- read.csv(paste("./weather/",files[j],sep=""))
      message("read")
      weather_data <- gtools::smartbind(weather_data,weather_data_j)
      rm(weather_data_j)
    }
  }


  

  weather_data_station <- weather_data %>% group_by(STATION,DATE) %>%
  summarise(TAVG=median(TAVG,na.rm = TRUE), #avg temp
            TMIN=median(TMIN,na.rm = TRUE), #min temp
            TMAX=median(TMAX,na.rm = TRUE), #mav temp
            PRCP=median(PRCP,na.rm = TRUE), #precipitation
            SNOW=median(SNOW,na.rm = TRUE), #snowfall
            AWND=median(AWND,na.rm = TRUE), #avg wind speed
            WSF2=median(WSF2,na.rm = TRUE),#fastest 2 min gust
            WDMV=mean(WDMV,na.rm = TRUE), #total wind movement
            WT01=sum(WT01,na.rm = TRUE), #WT01 - Fog, ice fog, or freezing fog (may include heavy fog)
            WT02=sum(WT02,na.rm = TRUE), #WT02 - Heavy fog or heaving freezing fog 
            WT03=sum(WT03,na.rm = TRUE), #WT03 - Thunder
            WT04=sum(WT04,na.rm = TRUE),  #WT04 - Ice pellets, sleet, snow pellets, or small hail
            WT05=sum(WT05,na.rm = TRUE), #WT05 - Hail (may include small hail)
            WT06=sum(WT06,na.rm = TRUE), #WT06 - Glaze or rime
            WT08=sum(WT08,na.rm = TRUE), #WT08 - Smoke or haze
            WT09=sum(as.numeric(WT09),na.rm = TRUE), #WT09 - Blowing or drifting snow
            WT11=sum(WT11,na.rm = TRUE), #WT11 - High or damaging winds
            )  %>%  mutate(Date=as_date(parse_date_time(DATE, orders = "Y-m-d"))) %>% select(-DATE)
  
    weather_data <- weather_data %>% group_by(DATE) %>%
    summarise(TAVG=median(TAVG,na.rm = TRUE), #avg temp
            TMIN=median(TMIN,na.rm = TRUE), #min temp
            TMAX=median(TMAX,na.rm = TRUE), #mav temp
            PRCP=median(PRCP,na.rm = TRUE), #precipitation
            SNOW=median(SNOW,na.rm = TRUE), #snowfall
            AWND=median(AWND,na.rm = TRUE), #avg wind speed
            WSF2=median(WSF2,na.rm = TRUE),#fastest 2 min gust
            WDMV=mean(WDMV,na.rm = TRUE), #total wind movement
            WT01=sum(WT01,na.rm = TRUE), #WT01 - Fog, ice fog, or freezing fog (may include heavy fog)
            WT02=sum(WT02,na.rm = TRUE), #WT02 - Heavy fog or heaving freezing fog 
            WT03=sum(WT03,na.rm = TRUE), #WT03 - Thunder
            WT04=sum(WT04,na.rm = TRUE),  #WT04 - Ice pellets, sleet, snow pellets, or small hail
            WT05=sum(WT05,na.rm = TRUE), #WT05 - Hail (may include small hail)
            WT06=sum(WT06,na.rm = TRUE), #WT06 - Glaze or rime
            WT08=sum(WT08,na.rm = TRUE), #WT08 - Smoke or haze
            WT09=sum(as.numeric(WT09),na.rm = TRUE), #WT09 - Blowing or drifting snow
            WT11=sum(WT11,na.rm = TRUE), #WT11 - High or damaging winds
            )  %>%  mutate(Date=as_date(parse_date_time(DATE, orders = "Y-m-d"))) %>% select(-DATE)
  
  saveRDS(weather_data_station,"weather_data_station.RDS")  
  saveRDS(weather_data,"weather_data.RDS")
}



#No need to rds ..local small file, manually created from google restuls
public_holidays <- read_csv("public_holidays.csv")

public_holidays <- public_holidays %>% mutate(Date=as_date(parse_date_time(Date, orders = "d b Y")))


if(file.exists("bike_stations.RDS")){
  bike_stations <- readRDS("bike_stations.RDS")
}else{

  bike_stations <- rgdal::readOGR("https://opendata.arcgis.com/datasets/a1f7acf65795451d89f0a38565a975b3_5.geojson")
  saveRDS(bike_stations,"bike_stations.RDS")
}

#rm(j,files)


```

## Introduction and Purpose
:::::: {.columns}
::: {.column width="60%"}

```{r, echo=FALSE, include=FALSE}
station_daily <- trip_data %>%
  mutate(Date=as_date(`Start date`),
         `Start station number`=as.numeric(`Start station number`)) %>% 
  group_by(Date,`Start station number`,`End station number`) %>%
  summarise(n=n()) %>% ungroup() 


station_demand <- station_daily  %>%  mutate(Station=`Start station number`) %>%
  group_by(Date,Station) %>%
  summarise(n=sum(n)) %>% ungroup() %>%
  mutate(season = metR::season(lubridate::month(Date))) %>%
  left_join(seasons, by="season") %>% select(-season) %>%
  left_join(public_holidays,by="Date") %>%
  mutate(weekend=if_else((lubridate::wday(Date,label=TRUE) %in% c("Sat","Sun")),TRUE,FALSE),
         season=season_name,
         public_holiday=if_else(is.na(Event),FALSE,TRUE,missing=FALSE),
         free_day = ifelse(weekend |public_holiday,"Non-working day","Working day")
         ) %>%
  group_by(free_day,season,Station) %>%
  summarise(daily_avg=mean(n)) %>% 
  mutate(Type="Start") %>%
  arrange(-daily_avg) %>% ungroup() %>%
  left_join((as.data.frame(bike_stations) %>% 
              mutate(TERMINAL_NUMBER=as.numeric(TERMINAL_NUMBER)) %>%
              select(Station=TERMINAL_NUMBER,LATITUDE,LONGITUDE)),
           by="Station")


workshop<- station_daily %>% filter(`Start station number`==0) %>% 
  group_by(Date) %>%
  summarise(In=sum(n)) %>%
  left_join( station_daily %>% filter(`End station number`==0) %>% 
  group_by(Date) %>%
  summarise(Out=sum(n)), by="Date")


stat <- trip_data %>% mutate(year=lubridate::year(`Start date`)) %>% filter(year==2017)
trips_stat <- round(nrow(stat)/10^6,2)
bike_fleet <- length(stat %>% pull(`Bike number`) %>% unique(.))

```


As the Council-owned bike sharing scheme for the City, *BikeShare* has the mission to provide a reliable, cost-effective bicycle sharing service across the Metropolitan area.  In 2017 alone, *BikeShare* served over 
`r sprintf("%.1f %%", trips_stat)` million trips with a fleet of `r formatC(bike_fleet, format="f", big.mark=",", digits=0)` bicycles.

As of today, bike repairs only occur when a bike is detected broken. Apart from the obvious impact on customer satisfaction when a bike breaks, this model is also inefficient from a workshop’s workload perspective, affecting workload and time to repair.

In order to address, this BikeShare can tap into the existing operational data and use analytics to optimise this problem.


:::
::: {.column width="40%"}

```{r traffic, echo=FALSE,include=FALSE}



fill.breaks <- quantile(station_demand$daily_avg,seq(0,1,0.2)) 
fill.colors <- RColorBrewer::brewer.pal(length(fill.breaks),"YlOrRd")


station_demand$`Daily Avg Traffic` <- cut(station_demand$daily_avg, fill.breaks)

station_demand <- station_demand %>% filter(!is.na(`Daily Avg Traffic`) &
                                            !is.na(LONGITUDE) &
                                              !is.na(LATITUDE))
                                        

bbox <- make_bbox(LONGITUDE, LATITUDE, station_demand, f = 0.005)
m <- get_map(bbox, source = "stamen", zoom = 13)

map <- ggmap(m)  +
  geom_point(data=station_demand,aes(x = LONGITUDE, y = LATITUDE,colour=`Daily Avg Traffic`)) +
  labs(title="Daily trips per Station") +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  scale_color_manual(values = fill.colors,
                         breaks =  waiver())  +
  facet_wrap(~season)

rm(m,bbox,fill.colors,fill.breaks)

```

```{r map, dpi = 800, dev.args = list(bg = 'transparent')}
map
```

:::
::::::

## Fleet maintenance volumes and windows need to be carefully chosen
:::::: {.columns}
::: {.column width="40%"}

In order to **<span style='color: brown;'>provide a good service</span>** for the users of the bike sharing scheme, **<span style='color: brown;'>unavailability needs to be avoided</span>**. Unavailability is caused by lack of maintenance or lack of supply. Thus, maintenance windows need to be selected **<span style='color: brown;'>smartly</span>**.

- Shared bikes are **<span style='color: brown;'>used every day of the week</span>**. Registered users provide a constant demand baseline.
- There is **<span style='color: brown;'>some seasonality</span>** in the demand: registered users brave through the winter keeping the numbers high.
- At first look, it seems that **<span style='color: brown;'>weather conditions are the biggest deterrent</span>** for bike usage. If it is too cold, or too hot or too windy casual users won't ride; registered users' numbers will also drop.
- Bikes will **<span style='color: brown;'>keep breaking</span>** - even if they are well built! Thus, **<span style='color: brown;'>every opportunity</span>** to undertake maintenance **<span style='color: brown;'>needs to be taken</span>**.
:::
::: {.column width="60%"}

```{r problem_statement_chart,dpi = 800, dev.args = list(bg = 'transparent'), echo = FALSE}

text_value <-   "**<span style='color: brown;'>Challenge</span>**: Using data analytics, we can leverage the **<span style='color: brown;'>weather forecast </span>** to optimise the fleet **<span style='color: brown;'>maintenace without compromising availibilty SLAs</span>**."

 daily_demand <- trip_data %>% mutate(Date =as_date(`Start date`)) %>% 
  group_by(Date,`Member type`) %>% summarise(n=n()) %>%
  pivot_wider(id_cols = Date,names_from = `Member type`,values_from = n) %>%
   left_join(public_holidays,by="Date") %>%
  mutate(Total=Casual+Member,
         day=lubridate::wday(Date,label=TRUE),
         weekend=if_else((day %in% c("Sat","Sun")),TRUE,FALSE),
         season = seasons[which(metR::season(month(Date))==seasons$season),]$season_name,
         public_holiday=if_else(is.na(Event),FALSE,TRUE,missing=FALSE)
         ) %>%
  left_join(weather_data,by="Date") %>%
  mutate(too_hot=(TMAX>=35 | TAVG>=30),
         too_cold=(TMIN<=-5 | TAVG<=4),
         too_windy=(AWND>30 |WSF2 >=50),
         too_rainy=(PRCP>15 |SNOW >10),
         too_bad = if_else( WT01>4 |
                     WT02>0 |
                     WT03>2 |
                     WT04>4 |
                     WT05>0 |
                     WT06>0 | 
                     WT08>5 |
                     WT09>3 |
                     WT11>0, TRUE,FALSE),
         Good_Weather=ifelse(!too_hot & !too_cold & !too_windy & !too_rainy & !too_bad,'Good','Bad')
  ) %>% pivot_longer(cols=c("Casual","Member"),names_to = "type", values_to = "trips") %>% select(-Total)

daily_demand$Good_Weather <-  factor(daily_demand$Good_Weather, levels=c('Good','Bad'))
daily_demand$type <-  factor(daily_demand$type, levels=c('Casual','Member'))
daily_demand$season <-  factor(daily_demand$season, levels=c('winter','spring','summer','autumn'))


g <-daily_demand %>%
  ggplot(aes(x=day,y=trips,color=Good_Weather, group=1)) +
  geom_jitter(shape=16, position=position_jitter(0.2)) +
  facet_grid(type ~ season, space='free_y') +
 theme_fivethirtyeight() + 
  scale_shape_tableau() +
  scale_colour_tableau('Classic Color Blind') +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "bottom",
        plot.title = element_text(size=14),
        axis.title.x = element_text(size = 10),
        axis.text.x = element_text(angle = 90, hjust = 1,size = 8),
        axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 8, angle = 90),
        legend.title=element_text(size=8),
        legend.text=element_text(size=8)) +
  labs(title="Bike Demand per Season and User Type",
       x="Day of the Week",
       y="Bikes Rented") +
  guides(col = guide_legend(ncol = 2),fill = guide_legend(title = "Weather", title.position = "left"))


t<-  textbox_grob(
  text_value,
  gp = gpar(fontsize = 12),
  box_gp = gpar(col = "grey80", fill = NA,lineend="round",lwd=3),
  r = unit(10, "pt"),
  padding = unit(c(10, 10, 10, 10), "pt"),
  width = unit(1, "npc"),
#  height  = unit(1, "npc"),
  margin = unit(c(10, 10,10, 10), "pt")
)

layout <- "
AAAA
AAAA
AAAA
AAAA
AAAA
####
CCCC
CCCC
"
g + t + plot_layout(design = layout) 
rm(g,t,layout,text_value)

```
:::
::::::

## By addressing uneven bike utilisation, it is possible to extend the life of the fleet
:::::: {.columns}
::: {.column width="40%"}

Some bicycles in the fleet are heavily used almost every single days, while others are seldomly ridden. As a result:

 - **<span style='color: brown;'>Heavily used bikes will be at higher risk of failure</span>** and accelerated asset ageing.
 - **<span style='color: brown;'>Unused bikes will be at risk of neglect</span>** - e.g. left with flat tyres in an low traffic station.
 
By obtaining a detailed log of all bike travel, it will be possible to identify which bikes are being used the most and calculate their risk of failing. This will allow to proactively book them for service.  

When those bike are taken out for service, they can be replaced with lower used bikes , evening asset utilisation.


:::
::: {.column width="60%"}

```{r bike usage, dpi = 800, dev.args = list(bg = 'transparent')}
t<- trip_data %>% mutate(Date=as_date(`Start date`)) %>%
  group_by(Date,`Bike number`) %>%
  summarise(trips=n()) %>% ungroup()
  
t<- t %>% mutate(Month=format(as.Date(Date), "%Y-%m")) %>%
  filter(Month %in% c("2017-05","2017-06","2017-07","2017-08"))

selection <- t %>% group_by(`Bike number`) %>% summarise(n=n()) %>% ungroup() %>% arrange(-n)


selection2 <- rbind(rbind(selection[1:7,],
                   selection[(nrow(selection)-6):nrow(selection),]),
                   selection[(round(nrow(selection)/2,0)-3):(round(nrow(selection)/2,0)+3),])
                   

bike_use <- t %>% filter(`Bike number` %in% selection2$`Bike number`) %>%
  ggplot(aes(Date,`Bike number` , fill= trips)) + 
  geom_tile()+
  scale_fill_gradient(low="lightblue", high="navy") +
   theme_fivethirtyeight() +
  labs(title="Daily trips - selected bikes from May to Aug 2017",
       x="Date",
       y="Bike Number") +
  theme(legend.position = "bottom",
        plot.title = element_text(size=12),
        axis.title.x = element_text(size = 10),
        axis.text.x = element_text(angle = 90, hjust = 1,size = 8),
        axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 8, angle = 90),
        legend.title=element_text(size=8),
        legend.text=element_text(size=8)) 


rm(t,selection,selection2)

bike_use
```

:::
::::::

## Data Collection to Optimise Maintenance workload and bike availability
:::::: {.columns}
::: {.column width="40%"}

To achieve the objectives, the below data needs to be collected:

- [Detailed trip data, itemised per each trip taken by each bicycle](https://www.capitalbikeshare.com/system-data)
- [Weather Observations/Forecast for each day of the year](https://www.ncdc.noaa.gov/cdo-web/datasets)
- Bike failure rates.
- Workshop's repair turnaround stats.

:::
::: {.column width="40%"}

Based on the data, the below metrics will be generated:


```{r table}

table_data <- data.frame(Metric=c("Daily Service Capacity",
                             "Need to Service Factor"),
                Description=c("Number of bikes that can be maintain any day, based on workshop capacity and predicted demand",
                              "Indicator whether a bike needs service ahead of probable failure"),
                     Source=c("Workshop Turnaround stats, Summary of daily trip data, weather",
                              "Detailed trip data")
                     )

table <- table_data %>% flextable() %>%
  width(j=1,width=1.5) %>%
  width(j=2:3,width=3) %>%
  bold(i = NULL, j = 1, bold = TRUE, part = "body") %>%
  bold(i = NULL, j = NULL, bold = TRUE, part = "header") %>%
  align(i = NULL, j = NULL, align = "left", part = "all") %>%
  border(
  i = NULL,
  j = NULL,
  border = NULL,
  border.top = NULL,
  border.bottom = fp_border(color = "grey"),
  border.left = NULL,
  border.right = NULL,
  part = "body"
)

knitr::knit_print(table)

```

These metrics (combined with spontaneous failure rates) should serve as input for a proactive service plan.

:::
::::::


## Next Steps
:::::: {.columns}
::: {.column width="40%"}

The next steps on this journey are:

- Source all missing data.
- Collect all data.
- Conduct analysis and try different models.
- Present results.

:::
::: {.column width="60%"}

```{r image2, out.width='100%'}
# Source : https://pixabay.com/images/id-4196725/
image_1 <- image_read('bike-sharing-4196725_1920.jpg')
plot(image_1)
```

:::
::::::


## Results and Deliverable
:::::: {.columns}
::: {.column width="40%"}

The generated metrics will be used to generated a day-by-day maintenance plan, by matching the optimal number of bicycle to service every day with a suggestion of which assets need to be service each day. This plan will include an allowance for unexpected service needs. 
This model will require weekly updates.

:::
::: {.column width="60%"}

```{r image3, out.width='100%'}
# Source : https://pixabay.com/images/id-659778/
image_1 <- image_read('cycling-659778_1920.jpg')
plot(image_1)
```

:::
::::::